---
title: "ST558 Project 1"
author: "David Arthur"
date: "6/14/2021"
output:
  github_document:
    toc: true
always_allow_html: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hablar)
library(tidyverse)
library(httr)
library(jsonlite)
library(knitr)
library(DT)
library(XML)
library(xml2)
library(lubridate)
```

# Function to get Team ID when name is input
```{r}
franchiseInfo <- GET("https://records.nhl.com/site/api/franchise") %>% 
  content("text") %>% 
  fromJSON(flatten = TRUE)

teamNameIdMatch <- franchiseInfo$data %>% select(id, mostRecentTeamId, fullName)

getTeamId <- function(team){
  Id <- teamNameIdMatch %>% filter(str_detect(fullName, team)) %>%
    select(id)
  if (nrow(Id) == 1){
    return(as.numeric(Id))
  } else {
    stop("try again entering team name or ID")
  }
}
```

## Records API functions
```{r}
recordsBaseURL <- "https://records.nhl.com/site/api"

replaceStrHTML <- function(x){
  for(i in 1:length(x)){
    if(!is.na(x[i])){
      x[i] <- x[i] %>%
        read_xml(as_html = TRUE) %>%
        xml_text() %>%
        str_replace_all(c("\r\n\t" = ";  ", "\r\n" = ""))
    }
  }
return(x)
}

getFranchiseInfo <- function(team = NULL){
  if(is.null(team)){
    fullURL <- paste0(recordsBaseURL, "/franchise")
    franchiseInfo <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  } else{
    if(!is.numeric(team)){team <- getTeamId(team)}
    fullURL <- paste0(recordsBaseURL, "/franchise?cayenneExp=id=", team)
    franchiseInfo <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  }
  return(as_tibble(franchiseInfo$data))
}

getTeamTotals <- function(team = NULL){
  if(is.null(team)){
    fullURL <- paste0(recordsBaseURL, "/franchise-team-totals")
    teamTotals <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  } else{
    if(!is.numeric(team)){team <- getTeamId(team)}
    fullURL <- paste0(recordsBaseURL, "/franchise-team-totals?cayenneExp=franchiseId=", team)
    teamTotals <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  }
  teamTotals <- teamTotals$data %>% relocate(teamId, teamName) %>%
    convert(fct(activeFranchise))
  return(as_tibble(teamTotals))
}

getSeasonRecords <- function(team = NULL){
  if(is.null(team)){
    fullURL <- paste0(recordsBaseURL, "/franchise-season-records")
    seasonRecords <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  } else{
  if(!is.numeric(team)){team <- getTeamId(team)}
  fullURL <- paste0(recordsBaseURL, "/franchise-season-records?cayenneExp=franchiseId=", team)
  seasonRecords <- GET(fullURL) %>% content("text") %>% fromJSON(flatten = TRUE)
  }
  seasonRecords <- seasonRecords$data %>%
    relocate(starts_with("franchise"), .after = id)
  return(as_tibble(seasonRecords))
}

getGoalieRecords <- function(team = NULL){
  if(is.null(team)){
    fullURL <- paste0(recordsBaseURL, "/franchise-goalie-records")
    goalieRecords <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  } else{
  if(!is.numeric(team)){team <- getTeamId(team)}
  fullURL <- paste0(recordsBaseURL, "/franchise-goalie-records?cayenneExp=franchiseId=", team)
  goalieRecords <- GET(fullURL) %>% content("text") %>% fromJSON(flatten = TRUE)
  }
  goalieRecords <- goalieRecords$data %>%
    relocate(firstName, lastName)
  return(as_tibble(goalieRecords))
}

getSkaterRecords<- function(team = NULL){
  if(is.null(team)){
    fullURL <- paste0(recordsBaseURL, "/franchise-skater-records")
    skaterRecords <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  } else{
  if(!is.numeric(team)){team <- getTeamId(team)}
  fullURL <- paste0(recordsBaseURL, "/franchise-skater-records?cayenneExp=franchiseId=", team)
  skaterRecords <- GET(fullURL) %>% content("text") %>% fromJSON(flatten = TRUE)
  }
  skaterRecords <- skaterRecords$data %>%
    relocate(firstName, lastName)
  return(as_tibble(skaterRecords))
}

getFranchiseDetail <- function(team = NULL){
  if(is.null(team)){
    fullURL <- paste0(recordsBaseURL, "/franchise-detail")
    franchiseDetail <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE)
  } else{
  if(!is.numeric(team)){team <- getTeamId(team)}
  fullURL <- paste0(recordsBaseURL, "/franchise-detail?cayenneExp=id=", team)
  franchiseDetail <- GET(fullURL) %>% content("text") %>% fromJSON(flatten = TRUE)
  }
  columnsHTML <- c("captainHistory", "coachingHistory", "generalManagerHistory", "retiredNumbersSummary")
  for(i in columnsHTML){
    franchiseDetail$data[, i] <- replaceStrHTML(franchiseDetail$data[, i])
  }
  franchiseDetail <- franchiseDetail$data %>%
    mutate(heroImageUrl = str_replace(heroImageUrl, " ", "%20")) %>%
    relocate(teamFullName) %>%
    mutate(dateAwarded = ymd(str_split(dateAwarded, "T", simplify = TRUE)[ ,1]))
  return(as_tibble(franchiseDetail))
}



detail <- getFranchiseDetail()
detail
franchise <- getFranchiseInfo()
totals <- getTeamTotals()
records <- getSeasonRecords()
# getSeasonRecords("Carolina")
goalie <- getGoalieRecords()
# getGoalieRecords("Hurricanes")
skater <- getSkaterRecords()
# getSkaterRecords(26)
kable(detail)
kable(franchise)
kable(totals)
kable(records)
kable(goalie[sample(1:nrow(goalie), 10), ])
kable(skater[sample(1:nrow(skater), 10), ])


```


##Stats API function
```{r Stats API}
getTeamIdStats <- function(team){
  Id <- teamNameIdMatch %>% filter(str_detect(fullName, team)) %>%
    select(mostRecentTeamId)
  if (nrow(Id) == 1){
    return(as.numeric(Id))
  } else {
    stop("try again entering team name or ID")
  }
}

getStats <- function(team = NULL){
  if(is.null(team)){
    teamStats <- GET("https://statsapi.web.nhl.com/api/v1/teams?expand=team.stats") %>%
      content("text") %>%
      fromJSON(flatten = TRUE) %>%
      as.data.frame()
    allTeamsStats <- allTeamsRanks <- data.frame()
    for (i in seq_along(teamStats$teams.id)){
      allTeamsStats <- bind_rows(allTeamsStats, teamStats[[9]][[i]][[1]][[1]][1, ])
      allTeamsRanks <- bind_rows(allTeamsRanks, teamStats[[9]][[i]][[1]][[1]][2, ])
    }
    allTeamsStats <- allTeamsStats %>% relocate(team.id, team.name) %>%
      rename_with(~ str_replace_all(.x, c("stat." = "", "team." = ""))) %>%
      select(-c(ends_with("Rank"), penaltyKillOpportunities)) %>%
      convert(num(wins:savePctg))
    allTeamsRanks <- allTeamsRanks %>% relocate(team.id, team.name) %>%
      rename_with(~ str_replace_all(.x, c("stat." = "", "team." = ""))) %>%
      select(-c(name, gamesPlayed, shootingPctg, savePctg, link)) %>%
      rename_with(~ paste0(.x, "Rank"), .cols = (wins: penaltyKillOpportunities))
    allTeams <- inner_join(allTeamsStats, allTeamsRanks, by = "id") %>%
      select(sort(names(.))) %>%
      relocate(id, name, wins, winsRank) %>%
      relocate(powerPlayGoalsRank, .after = powerPlayGoals) %>%
      relocate(link, .after = last_col())
    return(allTeams)
  } else{
    if(!is.numeric(team)) {team <- getTeamIdStats(team)}
    fullURL <- paste0("https://statsapi.web.nhl.com/api/v1/teams", "/", team, "?expand=team.stats")
    teamStats <- GET(fullURL) %>%
      content("text") %>%
      fromJSON(flatten = TRUE) %>%
      as.data.frame()
    oneTeamStats <- teamStats[[9]][[1]][[1]][[1]][1, ]
    oneTeamRanks <- teamStats[[9]][[1]][[1]][[1]][2, ]
    oneTeamStats <- oneTeamStats %>% relocate(team.id, team.name) %>%
      rename_with(~ str_replace_all(.x, c("stat." = "", "team." = ""))) %>%
      select(-c(ends_with("Rank"), penaltyKillOpportunities)) %>%
      convert(num(wins:savePctg))
    oneTeamRanks <- oneTeamRanks %>% relocate(team.id, team.name) %>%
      rename_with(~ str_replace_all(.x, c("stat." = "", "team." = ""))) %>%
      select(-c(name, gamesPlayed, shootingPctg, savePctg, link)) %>%
      rename_with(~ paste0(.x, "Rank"), .cols = (wins: penaltyKillOpportunities))
    oneTeam <- inner_join(oneTeamStats, oneTeamRanks, by = "id") %>%
      select(sort(names(.))) %>%
      relocate(id, name, wins, winsRank) %>%
      relocate(powerPlayGoalsRank, .after = powerPlayGoals) %>%
      relocate(link, .after = last_col())
    return(as_tibble(oneTeam))
  }
}
stats <- getStats()
kable(stats)
stats

# statsTibble <- as_tibble(stats)
# statsTibble %>% slice(seq(from = 1, to = nrow(statsTibble), by = 2)) %>% rename_with(~ str_replace_all(.x, c("stat." = "", "team." = "")))
#for stats without rankings
#str(testStats2[[9]][[11]][[1]][[1]][1,]) %>%
#  relocate(starts_with("team"))
```

## Wrapper function
```{r}
getChoice <- function(what, team = NULL){
  switch(what,
         franchise = getFranchiseInfo(team),
         totals = getTeamTotals(team),
         season = getSeasonRecords(team),
         goalie = getGoalieRecords(team),
         skater = getSkaterRecords(team),
         detail = getFranchiseDetail(team),
         stats = getStats(team),
         stop("try again entering what you want")
         )
}

getChoice("franchise", 26)
getChoice("totals")
getChoice("season", "Caro")
getChoice("goalie", "Rangers")
getChoice("skater", "Philadelphia Q")
getChoice("detail")
getChoice("stats", "Detroit")
kable(totals)
```

## EDA
Create new variables: Win/Loss Percentage (regular season only, ties excluded) and ratio of Goals For to Goals Against.  Scatter plot
```{r}
totalsNew <- totals %>% filter(gameTypeId == 2) %>%
  mutate(WLPctg = wins/(wins + losses)) %>%
  mutate(WLPctgCat = cut(WLPctg, c(0, 0.5, 1.0))) %>%
  mutate(goalsRatio = goalsFor / goalsAgainst) %>%
  mutate(goalsRatioCat = cut(goalsRatio, c(0, 1.0, 20))) %>%
  select(teamName, franchiseId, starts_with("WLPctg"), starts_with("goalsRatio"), activeFranchise)  
tableGoalsWL <- table(totalsNew$goalsRatioCat, totalsNew$WLPctgCat)

dimnames(tableGoalsWL)[[1]] <- c("Scored fewer total goals than gave up", "Scored more total goals than gave up")

kable(tableGoalsWL, col.names = c("Fewer wins than losses", "More wins than losses"), caption = "Count of teams with winning record vs. count of teams with losing record (all-time) by goals for/against ratio", row.names = TRUE)

gTotals <- ggplot(data = totalsNew)
gTotals + geom_point(aes(x = goalsRatio, y = WLPctg, col = activeFranchise)) +
  labs(title = "Scatterplot of goals ratio vs win percentage for each team (all-time)", x = "Ratio of goals scored to goals allowed (all seasons)", y = "Win percentage (ties excluded)") +
  scale_color_discrete(name = "", labels = c("non-active Franchise", "active Franchise"))
```

Create variables for count of retired jerseys and age of team, plus rate of jersey retirement = number retired / age of team

```{r}
detailReduced <- detail %>%
  select(id, teamFullName, retiredNumbersSummary, firstSeasonId) %>%
  mutate(numberRetiredJerseys = str_count(retiredNumbersSummary, "–")) %>%
  mutate(numberRetiredJerseys = ifelse(str_detect(retiredNumbersSummary, "In Honor of"), numberRetiredJerseys - 1, numberRetiredJerseys)) %>%
  mutate(numberRetiredJerseys = ifelse(is.na(numberRetiredJerseys), 0, numberRetiredJerseys)) %>%
  convert(chr(firstSeasonId)) %>%
  mutate(firstSeasonId = str_trunc(firstSeasonId, 4, ellipsis = "")) %>%
  convert(num(firstSeasonId)) %>%
  mutate(teamAge = 2021 - firstSeasonId) %>%
  mutate(jerseyRetirementRate = numberRetiredJerseys / teamAge) %>%
  mutate(jerseyRetirementRate = replace_na(jerseyRetirementRate, 0))

statsReduced <- stats %>%
  select(name, wins, losses, gamesPlayed, goalsPerGame, goalsAgainstPerGame) %>%
  mutate(seasonWLPctg = wins / (wins + losses)) %>%
  mutate(seasonWLPctgCat = cut(seasonWLPctg, c(0, 0.5, 1.0)))
  

totalsDetail <- left_join(totalsNew, detailReduced, by = c("teamName" = "teamFullName"))

totalsDetailStats <- inner_join(totalsDetail, statsReduced, by = c("teamName" = "name")) %>%
  mutate(ageCat = cut(teamAge, c(0, 25, 50, 75, 100, 125)))

tableAgeWL <- table(totalsDetailStats$ageCat, totalsDetailStats$seasonWLPctgCat)
dimnames(tableAgeWL)[[1]] <- c("0-25 years old", "26-50 years old", "51-75 years old", "76-100 years old", "101-125 years old")
names(dimnames(tableAgeWL)) <- c("Team Age", "")
kable(tableAgeWL, col.names = c("Fewer wins than losses", "More wins than losses"), caption = "Count of teams with winning record vs. count of teams with losing record for 2020-21 season by age of team", row.names = TRUE)

gWins <- ggplot(data = totalsDetailStats)
gWins + geom_bar(aes(x = ageCat, fill = seasonWLPctgCat), position = "dodge") +
  labs(title = "Active NHL Teams by age of franchise", x = "Team Age (years)", fill = "W/L ratio 2020-21")
```

```{r}
gWins + geom_boxplot(aes(x = seasonWLPctgCat, y = goalsPerGame)) +
  geom_point(aes(x = seasonWLPctgCat, y = goalsPerGame, col = ageCat), position = "jitter") +
  labs(title = "Boxplots of average goals scored per game", x = "2020-21 Win Percentage (ties excluded)", y = "Average goals scored per game", col = "Age of team") +
  scale_x_discrete(labels = c("More losses than wins", "More wins than losses"))
```

```{r}
gWins + geom_boxplot(aes(x = seasonWLPctgCat, y = goalsAgainstPerGame)) +
  geom_point(aes(x = seasonWLPctgCat, y = goalsAgainstPerGame, col = ageCat), position = "jitter") +
  labs(title = "Boxplots of average goals allowed per game", x = "2020-21 Win Percentage (ties excluded)", y = "Average goals allowed per game", col = "Age of team") +
  scale_x_discrete(labels = c("More losses than wins", "More wins than losses"))
```



```{r}
gWins + geom_histogram(aes(x = numberRetiredJerseys, fill = ageCat)) +
  labs(title = "Histogram of jerseys retired by team", x  = "Number of jerseys retired", fill = "Age of team")
```
```{r}
gWins + geom_point(aes(x = jerseyRetirementRate, y = WLPctg, col = ageCat)) +
  labs(title = "Scatterplot of jersey retirement rate vs win percentage (all seasons)", x = "Numbers retired divided by age of team", y = "All-time win percentage (ties excluded)", col = "Age of team") +
  scale_color_discrete(name = "Team age", labels = c("0-25 years", "26-50 years", "51-75 years", "76-100 years", "101-125 years"))
```

Numeric summaries
```{r}

```



